---
title: "12_factor_forcats"
author: "Doug"
date: '2022-08-18'
output: html_document
---

## Chapter 12: Factors with forcats

#### Introduction

-   factors are used to work with categorical variables
-   useful when you want to display character vectors in non alphabetical order
-   forcat is not part of the core tidyverse, load it explicitly

```{r}
library(tidyverse)
library(forcats)
```

#### Creating Factors

-   sorting categorical variables doesn't really work

```{r}
x1 <- c("Dec","Apr","Jan","Mar")
x2 <- c("Dec","Apr","Jam","Mar")
sort(x1)
```

-   creating a list displaying all possible values or valid levels
-   then create a factor, selecting values in x1 that correspond to month_levels
-   if values don't correspond you'd get an NA
-   using **parse_factor()** will show an error instead of NA

```{r}
month_levels <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
y1 <- factor(x1, levels = month_levels)
y2 <- factor(x2, levels = month_levels)
```

```{r}
y2 <- parse_factor(x2, levels = month_levels)
```

-   use **unique()** for having your levels match the original data order

```{r}
f1 <- factor(x1, levels = unique(x1))
f2 <- x1 %>% factor() %>% fct_inorder()
```

-   show list of levels with **levels()**

```{r}
levels(f2)
```

```{r}
gsss_cat
gss_cat %>% count(race)
```

```{r}
ggplot(gss_cat, aes(race)) +
  geom_bar()
```

-   levels that don't have values will automatically be dropped by ggplot2 but you can force them to be displayed

```{r}
ggplot(gss_cat, aes(race)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```

#### Modifying Factor Order

```{r}
relig <- gss_cat %>%
group_by(relig) %>%
summarize(
  age = mean(age, no.rm = TRUE),
  tvhours = mean(tvhours, na.rm = TRUE),
  n = n())
```

```{r}
ggplot(relig, aes(tvhours, relig))+
  geom_point()
```

-   the plot is difficult to interpret without any order to the display of values
-   **fct_reorder** will ask for a factor to modify levels, a numeric variable to reorder
-   so use it instead of the y variable by itself

```{r}
ggplot(relig, aes(tvhours, fct_reorder(relig, tvhours))) +
  geom_point()
```

-   **fct_reorder()** can be used outside of the aes, into a seperate mutate step

```{r}
relig %>%
  mutate(relig = fct_reorder(relig, tvhours)) %>%
  ggplot(aes(tvhours, relig)) +
  geom_point()
```

-   the problem here is that the categories are ordinal, so rearranging the order based on the values in age is not a good idea
-   so only use **fct_reorder()** with variables that are arbitrarily ordered

```{r}
rincome <- gss_cat %>%
  group_by(rincome) %>%
  summarize(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n())
```

```{r}
ggplot(rincome, aes(age, fct_reorder(rincome, age))) +
  geom_point()
```

-   however in these cases **fct_relevel()** can be used to bring some levels of a factor to the front of the display
-   control the placement of it with *after =* and *before =*

```{r}
ggplot(rincome, aes(age, fct_relevel(rincome, "Not applicable", "Don't know"))) +
  geom_point()
```

-   use **fct_reorder2()** to reorder the factor by the y values associated with the largest x values

```{r}
by_age <- gss_cat %>%
  filter(!is.na(age)) %>%
  group_by(age, marital) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n))
```

```{r}
ggplot(by_age, aes(age, prop, color = marital)) +
  geom_line(na.rm = TRUE)
```

```{r}
ggplot(by_age, aes(age, prop, color = fct_reorder2(marital, age, prop))) +
  geom_line() +
  labs(color = "marital")
```

-   use **fct_infreq()** for bar charts reordering levels in increasing frequency
-   often combined with **fct_rev()** to reverse the order

```{r}
gss_cat %>%
  mutate(marital = marital %>% fct_infreq() %>% fct_rev()) %>%
  ggplot(aes(marital)) +
  geom_bar()

gss_cat %>%
  mutate(marital = marital %>% fct_infreq() %>% fct_rev) %>%
  ggplot(aes(marital)) +
  geom_bar()
```

#### Modifying Factor Order

-   use **fct_recode()** to recode the values of each level, so you can combine multiple levels into one level
-   be careful using this as it could bring misleading results if categories are actually different

```{r}
gss_cat %>% count(partyid)

gss_cat %>%
  mutate(partyid = fct_recode(partyid,
                              "Republican, strong" = "Strong republican",
                              "Republican, weak" = "Not str republican",
                              "Independent, near rep" = "Ind,near rep",
                              "Independent, near dem" = "Ind,near dem",
                              "Democrat, weak" = "Not str democrat",
                              "Democrat, strong" = "Strong democrat")) %>%
  count(partyid)
```

```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
                              "Republican, strong" = "Strong republican",
                              "Republican, weak" = "Not str republican",
                              "Independent, near rep" = "Ind,near rep",
                              "Independent, near dem" = "Ind,near dem",
                              "Democrat, weak" = "Not str democrat",
                              "Democrat, strong" = "Strong democrat",
                              "Other" = "No answer",
                              "Other" = "Don't know",
                              "Other" = "Other party")) %>%
  count(partyid)
```

-   use **fct_collapse** for assigning a new name to multiple old names

```{r}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
                                other = c("No answer", "Don't know", "Other party"),
                                rep = c("Strong republican", "Not str republican"),
                                ind = c("Ind,near rep", "Independent", "Ind,near dem"),
                                dem = c("Not str democrat", "Strong democrat"))) %>%
  count(partyid)
```

-   lumping together all small groups for a simpler plot using **fct_lump()**
-   it defaults to lumping the smallest groups so the aggregate group is still the smallest group
-   the n parameter can be used to specify how many groups you're looking for

```{r}
gss_cat %>%
  mutate(relig = fct_lump(relig, n =10)) %>%
  count(relig, sort = TRUE) %>%
  print(n = Inf)
```

-   this took the 15 religion categories and lumped the bottom 5 into 'Other'
